---
title: "Aides Drug"
author: "Daniel Esteves"
date: "Summer 2024"
output: html_document
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
triUmp <-read.csv(file="triUmpsal.csv",header=TRUE)

druggies = select(triUmp, c(treatment, infected, DrugUser))
druggies = filter(druggies, DrugUser == "yes")
                  
```

# Introduction

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?.  If you skimp in this section expect to be marked down in your grade. Paint a picture of the project a couple of paragraphs is expected!-->

A new  preventative aides drug (triUmpsal) is under test in clinical trials. Consider only the DrugUsers, does this drug seem effective?  Hint - use dplyr to obtain that subpopulation with the filter command.

$H_0$: The drug triUmpsal has no effect on preventing AIDS among drug users

$H_A$: The drug triUmpsal is effective in preventing AIDS among drug users

# Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method. Discuss what data manipulations may be needed for example subsetting or other binning procedures and what sort of graphical displays and inferential tools will be used!-->


The variables 'treatment' and 'infected' are both categorical, meaning they sort data into different categories. The Cat~Cat helper file is appropriate to use, when analyzing them.  

# Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

## Descriptive Results

### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question. Bottom line is to explain the features of the graph that are relevant to conforming or departing to what is expected by your null hypothesis !-->

```{r}
barchartGC(~treatment + infected,data=druggies)
barchartGC(~infected + treatment,data=druggies, type="percent")

```

The chart shows that the group taking the drug had a lower percentage of people getting "aids" compared to the placebo group, while more people in the drug group experienced the "none" outcome.  

### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
table1 <- xtabs(~treatment + infected, data=druggies)
rowPerc(table1)
colPerc(table1)
```
The data shows that in the group taking the drug, only 2.89% got AIDS while 97.11% didn't, compared to the placebo group where 7.77% got AIDS and 92.23% didn't; overall, there were fewer AIDS cases in the drug group (24.29%) compared to the placebo group (75.71%).
## Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```

The p-value from the chi-squared test is smaller than 0.05, so we reject the null hypothesis, meaning there's a significant link between the treatment and the outcome, indicating the drug might help prevent AIDS.  

The Fisher's exact test gives a p-value of 0.0000000423, also less than 0.05, so we reject the null hypothesis again. The odds ratio is 0.3534:1, suggesting that the treatment was more effective in preventing AIDS than the placebo.  



# Discussion and Conclusion

<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->


The clinical trial data indicate that the drug TriUmpsal significantly reduces the number of AIDS cases among users compared to those who took a placebo. Both the chi-squared and Fisher's exact tests had p-values less than 0.05, showing strong evidence that the drug works. The odds ratio from Fisher's test, which was 0.3534:1, also suggested that TriUmpsal was more effective than the placebo in preventing AIDS, so we reject the null hypothesis.
